Reactive SINDy: Discovering governing reactions from concentration data

### 2018-12-06 docs/manuscript_jcp.tex
wip
describe results of mapk and pprey, update figure 7

We additionally demonstrate the algorithm
on time series data of the mitogen activated protein kinases (MAPK) pathway as 
an example for a bimodal system and on time series data of the
Lotka-Volterra system which describes predator-prey dynamics subject to
social friction. In both systems reactive SINDy recovers the generating 
reaction network whereas non-sparse estimation detects many false processes.

@@

Here we model the MAPK pathway with three stages of kinases $\mathrm{MAPK}$, $\mathrm{MAPKK}$, and $\mathrm{MAPKKK}$. The initial stimulus is called $\mathrm{S}$ and the final substrate to be activated is a transcription factor $\mathrm{TF}$. The ground truth reaction network reads
\begin{align*}
\text{Activations:} &\\
\mathrm{S}        + \mathrm{MAPKKK}  & \rightharpoonup \mathrm{S      } + \mathrm{MAPKKK*} \\
\mathrm{MAPKKK*}  + \mathrm{MAPKK }  & \rightharpoonup \mathrm{MAPKKK*} + \mathrm{MAPKK* } \\
\mathrm{MAPKK*}   + \mathrm{MAPK  }  & \rightharpoonup \mathrm{MAPKK* } + \mathrm{MAPK*  } \\
\mathrm{MAPK*}    + \mathrm{TF    }  & \rightharpoonup \mathrm{MAPK*  } + \mathrm{TF*    }  \\
\text{Deactivations:}&\\
\mathrm{MAPKKK*}    \mathrm{      }  & \rightharpoonup \mathrm{MAPKKK }  \\
\mathrm{MAPKK*}     \mathrm{      }  & \rightharpoonup \mathrm{MAPKK  }  \\
\mathrm{MAPK*}      \mathrm{      }  & \rightharpoonup \mathrm{MAPK   }  \\
\mathrm{TF*}        \mathrm{      }  & \rightharpoonup \mathrm{TF     }.
\end{align*}
For simplicity we assume phosphatase to be abundant such that deactivations 
effectively become first order reactions.
The external stimulus $\mathrm{S}$ is not consumed such that time integration of these 
reactions yields a steady state in which the concentration of response $[\mathrm{TF}]$ 
can be measured as a function of the stimulus concentration $[\mathrm{S}]$.
Using the rate constants given in Tab.~\ref{tab:reaction-library-mapk} we obtain the 
response curve given in Fig.~\ref{fig:mapk}a.

We generate concentration time series data of the MAPK reactions above at 
three different initial conditions, each differing in the amount of stimulus.
They are marked in Fig.~\ref{fig:mapk}a by vertical dashed lines.
These time series are concatenated to yield a dataset of 300 frames in total.
We use the library $\Theta$ of candidate reactions Tab.~\ref{tab:reaction-library-mapk}.
To determine the hyperparameter $\alpha=6.6\times 10^{-9}$ we shuffle the data and perform
15-fold cross validation. We obtain the estimated 
rate constants by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are given in Fig.~\ref{fig:mapk}b.
{\color{red}TODO describe}

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{figures_jcp/6.pdf}
    \caption{\label{fig:mapk}Application of reactive SINDy to the MAPK pathway system. \textbf{(a)}~The response curve of the MAPK cascade as a function of external stimulus given as a constant concentration $[\mathrm{S}]$. The activity is the steady state concentration of activated transcription factors $[\mathrm{TF*}]$. Dashed lines show the values of $[\mathrm{S}]$ at which concentration time series data was generated. \textbf{(b)}~Estimated rate coefficients of candidate reactions after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) and the ground truth model for comparison.}
\end{figure}

\subsection{Application to Lotka-Volterra system\label{sec:lv}}
As biological pathways often exhibit oscillatory behavior~\cite{Iqbal2008tnf} which can stem 
from positive or negative feedback loops~\cite{Shin2009positive} we apply reactive SINDy to
an idealized oscillatory system, namely the Lotka-Volterra system. 
The predator-prey dynamics of two species $\mathrm{A}$ (prey) and $\mathrm{B}$ 
(predator) is defined by the reactions
\begin{align*}
\mathrm{A}              & \rightharpoonup \mathrm{A} + \mathrm{A} &  & \text{(prey growth)},\\
\mathrm{A} + \mathrm{B} & \rightharpoonup \mathrm{B} + \mathrm{B} &  & \text{(predator eats prey)},\\
\mathrm{B}              & \rightharpoonup \emptyset &  & \text{(predator decay)},\\
\mathrm{A} + \mathrm{A} & \rightharpoonup \emptyset &  & \text{(prey friction)},\\
\mathrm{B} + \mathrm{B} & \rightharpoonup \emptyset &  & \text{(predator friction)}.
\end{align*}
From this model we generate concentration time series data with $200$ frames 
which is displayed in Fig.~\ref{fig:lv}a. We construct the library $\Theta$ 
of candidate reactions given in Tab.~\ref{tab:reaction-library-lv}. 
We determine the hyperparameter $\alpha=2.7\times10^{-7}$ by shuffling 
the data and performing 5-fold cross validation. We obtain the estimated 
rate constants by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are given in Fig.~\ref{fig:lv}b.
{\color{red}TODO describe}

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{figures_jcp/7.pdf}
    \caption{\label{fig:lv}Application of reactive SINDy to the Lotka-Volterra system with social friction. \textbf{(a)}~Concentration data as a function of time for predator and prey species. \textbf{(b)}~Estimated rate coefficients of candidate reactions after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) estimation and the ground truth model for comparison.}
\end{figure}

@@

In this work we have extended the SINDy method to reactive SINDy,
not only parsimoniously detecting potentially nonlinear terms in a
dynamical system from noisy data, but also yielding, in this case,
a sparse set of rates with respect to generating reactions (\ref{method:the-reactions}).
Mathematically this has been achieved by permitting vector-valued
basis functions and obtaining a tensor linear regression problem.
We have applied this method on data generated from a gene regulation
network and could successfully recover the underlying model.

The studies of Sec.~\ref{sec:case-2} and Sec.~\ref{sec:case-3}
have shown that the applied regularization terms can mitigate noise
up to a certain degree compared to the unregularized method, so that
identification of the reaction network is more robust and closer to
the ground truth.

Potentially, this method could be used to identify reaction networks
from time series measurements even if the initial conditions are not
always exactly identical, as was demonstrated in Sec.~\ref{sec:case-3}.
One apparent limitation is that the method can only be applied if
the data stems from the equilibration phase, as the concentration-based
approach has derivatives equal zero in the equilibrium, which precludes
the reaction dynamics to be recovered. In future work, we will consider
the identification of reaction schemes from instantaneous fluctuations
of particle numbers in equilibrium.

>>>>

We additionally demonstrate the algorithm
on time series data of the mitogen activated protein kinases (MAPK) pathway as 
an example for a bimodal system and on time series data of the
Lotka-Volterra system which describes oscillatory predator-prey dynamics subject to
social friction. In both systems reactive SINDy recovers the generating 
reaction network whereas non-sparse estimation detects many false processes.

@@

Here we model the MAPK pathway with three stages of kinases $\mathrm{MAPK}$, $\mathrm{MAPKK}$, and $\mathrm{MAPKKK}$. The initial stimulus is called $\mathrm{S}$ and the final substrate to be activated is a transcription factor $\mathrm{TF}$. The ground truth reaction network reads
\begin{align*}
\intertext{Activations:}
\mathrm{S}        + \mathrm{MAPKKK}  & \rightharpoonup \mathrm{S      } + \mathrm{MAPKKK*} \\
\mathrm{MAPKKK*}  + \mathrm{MAPKK }  & \rightharpoonup \mathrm{MAPKKK*} + \mathrm{MAPKK* } \\
\mathrm{MAPKK*}   + \mathrm{MAPK  }  & \rightharpoonup \mathrm{MAPKK* } + \mathrm{MAPK*  } \\
\mathrm{MAPK*}    + \mathrm{TF    }  & \rightharpoonup \mathrm{MAPK*  } + \mathrm{TF*    }
\intertext{Deactivations:}
\mathrm{MAPKKK*}    \mathrm{      }  & \rightharpoonup \mathrm{MAPKKK }  \\
\mathrm{MAPKK*}     \mathrm{      }  & \rightharpoonup \mathrm{MAPKK  }  \\
\mathrm{MAPK*}      \mathrm{      }  & \rightharpoonup \mathrm{MAPK   }  \\
\mathrm{TF*}        \mathrm{      }  & \rightharpoonup \mathrm{TF     }.
\end{align*}
For simplicity we assume phosphatase to be abundant such that deactivations 
effectively become first order reactions.
The external stimulus $\mathrm{S}$ is not consumed such that time integration of these 
reactions yields a steady state in which the response, i.e. the concentration $[\mathrm{TF*}]$ 
can be measured as a function of the stimulus concentration $[\mathrm{S}]$.
Using the rate constants given in Tab.~\ref{tab:reaction-library-mapk} we obtain the 
response curve given in Fig.~\ref{fig:mapk}a.

We generate concentration time series data of the MAPK reactions above at 
three different initial conditions, each differing in the amount of stimulus $[\mathrm{S}]$.
They are marked in Fig.~\ref{fig:mapk}a by vertical dashed lines.
These time series are concatenated to yield a dataset of 300 frames in total.
We use the library $\Theta$ of candidate reactions Tab.~\ref{tab:reaction-library-mapk}.
To determine the hyperparameter $\alpha=6.6\times 10^{-9}$ we shuffle the data and perform
15-fold cross validation. We obtain the estimated 
rate constants by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are given in Fig.~\ref{fig:mapk}b.
Least-squares estimation detects 5 of the 8 reaction processes that belong to the 
ground truth model. However it also detects 12 other reaction 
processes ($\theta_{18}$ - $\theta_{29}$) which are not in the ground truth.
Reactive SINDy estimation detects all reactions of the ground truth, 
two processes ($\theta_{4}$ and $\theta_{8}$) show deviations in 
rate constants. Generally reactive SINDy yields a sparse model which allows further
simplification of the reaction network by dropping out reaction processes that lie 
beneath a certain cutoff. In this case for example a cutoff of $0.25$ would directly recover
the ground truth reaction network.
Quantitatively, one may consider the $L_1$ norm of the relative distance 
of estimated rate constants $\hat{\xi}_r$ to the non-zero rate 
constants of the ground truth $\xi_r$
\begin{equation*}
\sum_{r=1}^{8} \left|(\hat{\xi}_{r} - \xi_r)/\xi_r\right|
\end{equation*}
which yields $167\%$ error for least-squares and $21\%$ error for reactive SINDy.

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{figures_jcp/6.pdf}
    \caption{\label{fig:mapk}Application of reactive SINDy to the MAPK pathway system. \textbf{(a)}~The response curve of the MAPK cascade as a function of external stimulus given as a constant concentration $[\mathrm{S}]$. The activity is the steady state concentration of activated transcription factors $[\mathrm{TF*}]$. Dashed lines show the values of $[\mathrm{S}]$ at which concentration time series data was generated. \textbf{(b)}~Estimated rate coefficients of candidate reactions (see Tab.~\ref{tab:reaction-library-mapk}) after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) and the ground truth model for comparison.}
\end{figure}

\subsection{Application to Lotka-Volterra system\label{sec:lv}}
As biological pathways often exhibit oscillatory behavior~\cite{Iqbal2008tnf} which can stem 
from positive or negative feedback loops~\cite{Shin2009positive} we apply reactive SINDy to
an idealized oscillatory system, namely the Lotka-Volterra system. 
The predator-prey dynamics of two species $\mathrm{A}$ (prey) and $\mathrm{B}$ 
(predator) is defined by the reactions
\begin{align*}
\mathrm{A}              & \rightharpoonup \mathrm{A} + \mathrm{A} &  & \text{(prey growth)},\\
\mathrm{A} + \mathrm{B} & \rightharpoonup \mathrm{B} + \mathrm{B} &  & \text{(predator eats prey)},\\
\mathrm{B}              & \rightharpoonup \emptyset &  & \text{(predator decay)},\\
\mathrm{A} + \mathrm{A} & \rightharpoonup \emptyset &  & \text{(prey friction)},\\
\mathrm{B} + \mathrm{B} & \rightharpoonup \emptyset &  & \text{(predator friction)}.
\end{align*}
From this model we generate concentration time series data with $200$ frames 
which is displayed in Fig.~\ref{fig:lv}a. We construct the library $\Theta$ 
of candidate reactions given in Tab.~\ref{tab:reaction-library-lv}. 
We determine the hyperparameter $\alpha=2.7\times10^{-7}$ by shuffling 
the data and performing 5-fold cross validation. We obtain the estimated 
rate constants by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are given in Fig.~\ref{fig:lv}b.
Least-squares estimation detects all reactions of the ground truth model but also
two false processes ($\theta_{6}$ and $\theta_{7}$) with a higher rate than
the two first true processes ($\theta_{1}$ and $\theta_{2}$). Reactive SINDy detects
the true reaction network with minor deviations in rate constants. As in 
Sec.~\ref{sec:mapk} we consider the $L_1$ norm of the relative distance to the ground truth
for non-zero rate constants
\begin{equation*}
\sum_{r=1}^{5} \left|(\hat{\xi}_{r} - \xi_r)/\xi_r\right|
\end{equation*}
which yields $75\%$ error for least-squares and $7\%$ error for reactive SINDy.

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{figures_jcp/7.pdf}
    \caption{\label{fig:lv}Application of reactive SINDy to the Lotka-Volterra system with social friction. \textbf{(a)}~Concentration data as a function of time for predator and prey species. \textbf{(b)}~Estimated rate coefficients of candidate reactions (see Tab.~\ref{tab:reaction-library-lv}) after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) estimation and the ground truth model for comparison.}
\end{figure}

@@

In this work we have extended the SINDy method to reactive SINDy,
not only parsimoniously detecting potentially nonlinear terms in a
dynamical system from noisy data, but also yielding, in this case,
a sparse set of rates with respect to generating reactions (\ref{method:the-reactions}).
Mathematically this has been achieved by permitting vector-valued
basis functions and obtaining a tensor linear regression problem.
We have applied this method on data generated from a gene regulation
network, a MAPK pathway, and a Lotka-Volterra system and 
could successfully recover the underlying reaction networks.

The studies of Sec.~\ref{sec:case-2} and Sec.~\ref{sec:case-3}
have shown that the applied regularization terms can mitigate noise
up to a certain degree compared to the unregularized method, so that
identification of the reaction network is more robust and closer to
the ground truth.
Potentially, this method could be used to identify reaction networks
from time series measurements even if the initial conditions are not
always exactly identical, as was demonstrated in Sec.~\ref{sec:case-3}.

One apparent limitation is that the method can only be applied if
the data stems from the equilibration phase, as the concentration-based
approach has derivatives equal zero in the equilibrium, which precludes
the reaction dynamics to be recovered. Thus, in the case of oscillatory 
systems the reaction network can be recovered robustly.

In future work, we will consider
the identification of reaction schemes from instantaneous fluctuations
of particle numbers in equilibrium.
